---
title: "5_Assignment 4"
output: html_document
date: "2026-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(readr.show_col_types = FALSE)

# Loading in the needed libraries
library(readr)
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```

# Group 5: Assignment 1

## 1. Study Objectives

### 1.1 Study Aim

We will be investigating the incidence of `Asthma_possible` and `Asthma_narrow`. To be able to do this, we will need the following:

#### 1.1.1 Including

-   Population:
    -   Age: 0-120
    -   Sex: Male / Female
-   Period: 2010-2022
-   Age Bands:
    -   0 - 19
    -   20 - 39
    -   40 - 59
    -   60+

#### 1.1.2 Excluding

-   Invalid sex
-   Missing person ID
-   Observation time is outside 2010-2022
-   Age is outside of the range

## 2. Clean the data

The following R code will load in the needed tables for us. Note, that we do not need to load the **VACCINES** data, or the **VISIT_OCCURANCE** data. As these will not help us in identifying the incidence rate for `Asthma_possible`, and `Asthma_narrow`.

```{r load_data}
# Loading in the data
persons <- read_csv("Data/PERSONS.csv")
observation_periods <- read_csv("Data/OBSERVATION_PERIODS.csv")
events <- read_csv("Data/EVENTS.csv")
asthma_codelist <- read_excel("Codelist/R_ASTHMA_COV.xlsx")
```

### 2.1 Identifing Key Variables

```{r}
# Checking the dimensions of the tables that we will be working with now
dim(persons)
dim(observation_periods)
dim(events)

print("---")

# Checking the names of the columns of the data
names(events)
print("---")
names(observation_periods)
print("---")
names(persons)
```

Here, we can see that both our persons and observation_periods table contain 500_000 entries, while events contain 126189 entries. The data models contain the following needed information:

-   **EVENTS**:
    -   `start_date_record` is in the events data
    -   `person_id` is in the events data
    -   `event_code` is in the events data
-   **OBSERVATION_PERIODS**:
    -   `person_id` is in the observation_periods data
    -   `op_start_date` is in the observation_periods data
    -   `op_end_date` is in the observation_periods data
-   **PERSONS**:
    -   `person_id` is in the persons data
    -   `day_of_birth` is in the persons data
    -   `month_of_birth` is in the persons data
    -   `year_of_birth` is in the persons data
    -   `sex_at_instance_creation` is in the persons data

### 2.2 Clean the data

In this section of the report we will be cleaning the data.

#### 2.2.1 Remove duplicates

Using the following code, we will be able to remove duplicates.

```{r}
# Check for duplicate persons in the persons table
sum(duplicated(persons$person_id))

# Check for duplicates in the observation periods table
sum(duplicated(observation_periods))
  
# Check for duplicates in the events table
sum(duplicated(events))
```

We can see that there are no duplicates for `person_id` in the persons table. There are also no duplicates in the observation_periods. However, in the events table there do seem to be 498 duplicate entries. Therefore, we remove those.

```{r}
events <- events %>% distinct()
```

#### 2.2.2 Handle Missing / Invalid Data

```{r}
# Checking if the person_id is missing from any of the persons, observation_periods, or events table
sum(is.na(persons$person_id))
sum(is.na(observation_periods$person_id))
sum(is.na(events$person_id))
sum(persons$person_id == "")
sum(observation_periods$person_id == "")
sum(events$person_id == "")

# Check if the persons table is missing values for sex_at_instance_creation, or year_of_birth
sum(is.na(persons$sex_at_instance_creation))
sum(is.na(persons$year_of_birth))
sum(persons$sex_at_instance_creation == "")
sum(persons$year_of_birth == "")
```

Currently, the `person_id` is nowhere missing in the persons, observation_periods and events table. Also, the `sex_at_instance_creation`, and `year_of_birth` is not missing from the persons data.

#### 2.2.3 Standardize date formats and check for valid entires

```{r}
# Check for invalid dates, where the observation end date is before the observation start date
sum(observation_periods$op_end_date < observation_periods$op_start_date)

# Easily check if we have birth years outside of the time requirements
summary(persons$year_of_birth)

# As the maximum is 2024, we want to check how many birth years are over 2022
sum(persons$year_of_birth > 2022)

# Check how many peoples death year is before their birth year
sum(
  !is.na(persons$year_of_death) &
  !is.na(persons$year_of_birth) &
  persons$year_of_death < persons$year_of_birth
)
```

We can see here that currently, 457 records from the observation_periods have an operation end_date before the operation start_date. Therefore, these must be excluded. We can also see that there are 10987 entries with a birth year after the end date of our observation period. We shouldn't include these, but these will be automatically excluded when creating the study population in the next step.

```{r}
# Excluding entries with operation end_date before the operation start_date from the observation_periods table
observation_periods <- observation_periods %>%
  filter(op_end_date >= op_start_date)
```

### 2.3 Create the study population

In this section of the report, we will be building the study population.

#### 2.3.1 Merge PERSONS with OBSERVATION_PERIODS

In this section of code, we will be merging the persons table, with the observation_periods table. We will also take a quick look at this table, so we can see these have correctly been merged.

```{r}
# Merging the persons and observation_periods table
study_population <- observation_periods %>%
  inner_join(persons, by = "person_id")

# Quickly checking what these merged data sources look like
dim(study_population)
head(study_population)
```

#### 2.3.2 Exclude subjects with impossible birthdates

In this section of code, we will be removing everyone from the study population in case they are born after 2022. If that is the case, they can't have any valid events.

```{r}
# We remove everyone from the population that is born after 2022. We know that the lower bound will not exclude anyone as the earliest birth year is 1934, which means the person is less than 120 years old, also we don't have to look at missing values as these do not occur in this data source
study_population <- study_population %>%
  filter(
    year_of_birth <= 2022
  )
```

#### 2.3.3 Impute missing `birth_date` and `death_date` values

If we find any missing values in the birth month / day, or the death month / day, we will automatically plug the 15th of July in those. Note, that this means that we are filling in a lot of incorrect values for when a person has died. However, these values cannot be converted to a proper date field in the next step. Thus `birth_date` and `death_date` will be filled correctly.

```{r}
# Make sure to replace the NA values from month_of_birth, day_of_birth, month_of_death, day_of_death, to be sure that the 15th of July is used.
study_population <- study_population %>%
  mutate(
    month_of_birth = replace_na(month_of_birth, "07"),
    day_of_birth   = replace_na(day_of_birth, "15"),
    month_of_death = replace_na(month_of_death, "07"),
    day_of_death  = replace_na(day_of_death, "15"),
  )
```

#### 2.3.4 Convert date variables from character strings to date format

We will now be converting the dates from strings, to a proper date format. This will be done using the `lubridate` library.

```{r}
# Make sure that we create proper date format characters for the birth and death date in the study_population
study_population <- study_population %>%
  mutate(
    birth_date = ymd(paste(year_of_birth, month_of_birth, day_of_birth, sep = "-")),
    death_date = ymd(paste(year_of_death, month_of_death, day_of_death, sep = "-"))
  )

# Make sure that the op_start_date and op_end_date are properly converted to the dates
study_population <- study_population %>%
  mutate(
    op_start_date = ymd(op_start_date),
    op_end_date   = ymd(op_end_date)
  )

# Make sure that start_date_record from events is properly converted to a date
events <- events %>%
  mutate(
    start_date_record = ymd(start_date_record)
  )

# Quickly show an overview of what these dates look like
summary(study_population$birth_date)
summary(study_population$death_date)
summary(study_population$op_start_date)
summary(events$start_date_record)
```

Note again, that the death_date contains a lot of values that are `NA` (also, see the error). This is because the year is missing for a lot of people, as they haven't died.

#### 2.3.5 Apply inclusion / exclusion criteria to observation start and end dates

As stated before, the observation period will be between 2010 and 2022. Therefore, we set the following start date: `2010-01-01`, and the following end date: `2022-12-31`. We will exclude all entries that contain an observation that is not within this time period. Next to that, we also remove entries where the end date happened before the start date.

```{r}
# Define the start and end dates of the observation
study_start <- as.Date("2010-01-01")
study_end   <- as.Date("2022-12-31")

# Exclude all entries with an observation end date before the start of the observation time period, or a start date outside of the observation time period
study_population <- study_population %>%
  filter(
    op_end_date >= study_start,
    op_start_date <= study_end
  )

# Exclude all entries where the observation end date is before the observation start date
study_population <- study_population %>%
  filter(op_end_date > op_start_date)

# As it is also stated that we will only be looking at "M", and "F", we will want to remove all other entries where the `sex_at_instance_creation` is not "M", or "F"
study_population <- study_population %>%
  filter(sex_at_instance_creation %in% c("M", "F"))
```

Last but not least, it is important to note here, that we have excluded all entries where `sex_at_instance_creation` is not either "M", or "F". As per the objective, we want to calculate the incidence of asthma in males and females.

### 2.4 Fix the code list for filtering

In this section, we will be fixing the code listing, such that it will be easier to filter these correctly.

#### 2.4.1 Ensure tags are stored and formatted correctly

```{r}
# Quickly show the column names and the first few entries of the table
names(asthma_codelist)
head(asthma_codelist)
```

As we want to study asthma, in both the possible and narrow case, we will need the following columns:

-   Coding system
-   Code
-   Tag

These are all available in the data source. Thus we can continue.

#### 2.4.2 Validate columns for coding system

We want to check whether there are any missing values from the Coding system and what coding systems exactly have been used. As there are no missing values, we won't have to remove any entries.

```{r}
# Check which coding systems are available in the data source
table(asthma_codelist$"Coding system")

# Check if the coding system is missing anywhere
sum(is.na(asthma_codelist$"Coding system"))
sum(asthma_codelist$"Coding system" == "")
sum(asthma_codelist$"Coding system" == "-")
```

#### 2.4.3 Address missing or empty codes

Now, we are going to check if there are any empty, or missing codes.

```{r}
# Check if there are any codes missing.
sum(is.na(asthma_codelist$Code))
sum(asthma_codelist$Code == "")
sum(asthma_codelist$Code == "-")
```

As there are a few missing, we will need to remove these from the table.

```{r}
# Remove entries from the asthma_codelist if the value in "Code" is "-"
asthma_codelist <- asthma_codelist %>%
  filter(`Code` != "-")
```

Looking through the table, we can see that `Code` from `asthma_codelist` no longer contains the value "-"

```{r}
asthma_codelist
```

## 3 Aggregate Person-Time

### 3.1 Determine Observation Periods

We defined follow-up from the latest entry (study start or observation start) to the earliest exit (event, death, or study end).

#### 3.1.1 Define start and end dates for follow-up

```{r}

# Extract specific codes based on tags for filtering events
narrow_codes <- asthma_codelist %>% filter(Tags == "Narrow") %>% pull(Code)
possible_codes <- asthma_codelist %>% filter(Tags %in% c("Narrow", "possible")) %>% pull(Code)

# Identify the first Asthma_narrow event date for each person to determine censoring
first_event_dt <- events %>%
  filter(event_code %in% narrow_codes) %>%
  group_by(person_id) %>%
  summarise(date_first_asthma = min(start_date_record, na.rm = TRUE)) %>%
  ungroup()

# Join the censoring dates back to the study population
study_population <- study_population %>%
  left_join(first_event_dt, by = "person_id")


```

#### 3.1.2 Censor data at the first event occurrence

```{r}
# Calculate the final follow-up dates using existing study_start and study_end
study_population <- study_population %>%
  mutate(
    # Start of follow-up
    fup_start = pmax(study_start, op_start_date),
    # End of follow-up (censoring)
    fup_end = pmin(study_end, op_end_date, death_date, date_first_asthma, na.rm = TRUE)
  ) %>%
  # Ensure follow-up duration is positive
  filter(fup_start < fup_end)




```

### 3.2 Stratify the data

We performed time-splitting to assign each individual’s follow-up to yearly slices and dynamic age bands.

#### 3.2.1 Define strata by year and age groups

```{r}
# Split follow-up into yearly records and assign age bands
py_stratified <- study_population %>%
  rowwise() %>%
  # Create a sequence of years covered by the follow-up
  mutate(year = list(year(fup_start):year(fup_end))) %>%
  unnest(year) %>%
  mutate(
    # Dates within the specific calendar year
    yr_start = pmax(fup_start, as.Date(paste0(year, "-01-01"))),
    yr_end   = pmin(fup_end, as.Date(paste0(year, "-12-31"))),
    # Calculate person-years for the year
    days = as.numeric(yr_end - yr_start) + 1,
    person_years = days / 365.25,
    # Assign dynamic age bands based on the year
    age_at_year = year - year_of_birth,
    age_band = case_when(
      age_at_year >= 0  & age_at_year <= 19 ~ "0-19",
      age_at_year >= 20 & age_at_year <= 39 ~ "20-39",
      age_at_year >= 40 & age_at_year <= 59 ~ "40-59",
      age_at_year >= 60 ~ "60+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_band))


```

### 3.3 Sum Person-Time

We aggregated total person-years by year, sex, and age band to create the study denominators.

#### 3.3.1 Calculate total person-time for each stratum

```{r}
# Aggregate denominators
denominator_data <- py_stratified %>%
  group_by(year, sex_at_instance_creation, age_band) %>%
  summarise(total_person_years = sum(person_years), .groups = "drop")

# Preview the aggregated person-years
head(denominator_data)


```

## 4 dentify and Count Events

### 4.1 Define the Event of Interest

We defined asthma events using the asthma codelist. Events were classified as Asthma_narrow or Asthma_possible based on the tags in the codelist. Only events occurring during the individual follow-up period were included.

#### 4.1.1 Specify what criteria qualifies as an event

# Identify asthma events and classify them as Narrow or Possible

```{r}
asthma_events <- events %>%
  filter(event_code %in% possible_codes) %>%
  mutate(
    asthma_type = case_when(
      event_code %in% narrow_codes ~ "Asthma_narrow",
      TRUE ~ "Asthma_possible"
    )
  )
```

### 4.2 Count Events per Stratum

Asthma events were stratified by calendar year, sex, and age band. Only one event per subject was counted within each stratum. These counts are going to be used as the numerators for the incidence rate calculations.

#### 4.2.1 Count only one event per subject per stratum

```{r}
# Merge asthma events with study population follow-up dates
asthma_events <- asthma_events %>%
  inner_join(
    study_population %>%
      select(person_id, fup_start, fup_end, birth_date, year_of_birth, sex_at_instance_creation),
    by = "person_id"
  ) %>%
  
  # Keep only events occurring during follow-up
  filter(
    start_date_record >= fup_start,
    start_date_record <= fup_end
  ) %>%
  
  # Define year and age band at event
  mutate(
    year = year(start_date_record),
    age_at_event = year - year_of_birth,
    age_band = case_when(
      age_at_event >= 0  & age_at_event <= 19 ~ "0-19",
      age_at_event >= 20 & age_at_event <= 39 ~ "20-39",
      age_at_event >= 40 & age_at_event <= 59 ~ "40-59",
      age_at_event >= 60 ~ "60+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_band)) %>%
  
  # Count only one event per subject per stratum
  distinct(person_id, year, age_band, asthma_type, .keep_all = TRUE)


# Aggregate numerator counts
numerator_data <- asthma_events %>%
  group_by(year, sex_at_instance_creation, age_band, asthma_type) %>%
  summarise(events = n(), .groups = "drop")

# Preview event counts
head(numerator_data)
```

## 5. Calculate Incidence Rates

### 5.1 Formula

For each stratum, incidence rate is calculated as: $$
IR = \frac{\text{events}}{\text{person-years}} \times 1000
$$

### 5.2 Calculate IR for Each Stratum

```{r incidence_by_stratum}
asthma_types <- c("Asthma_possible", "Asthma_narrow")

# Join numerators and denominators for every stratum
incidence_data <- denominator_data %>%
  crossing(asthma_type = asthma_types) %>%
  left_join(
    numerator_data,
    by = c("year", "sex_at_instance_creation", "age_band", "asthma_type")
  ) %>%
  mutate(
    events = replace_na(events, 0L),
    incidence_per_1000py = if_else(
      total_person_years > 0,
      (events / total_person_years) * 1000,
      NA_real_
    )
  ) %>%
  arrange(asthma_type, year, sex_at_instance_creation, age_band)

# Preview incidence by stratum
head(incidence_data)
```

### 5.3 Report Incidence Rates as Numbers

```{r incidence_tables_for_report}
# Full stratum-level table (year, sex, age band, definition)
incidence_table_stratum <- incidence_data %>%
  mutate(
    sex = recode(sex_at_instance_creation, "F" = "Female", "M" = "Male"),
    total_person_years = round(total_person_years, 2),
    incidence_per_1000py = round(incidence_per_1000py, 3)
  ) %>%
  select(
    year, sex, age_band, asthma_type, events,
    total_person_years, incidence_per_1000py
  ) %>%
  arrange(asthma_type, year, sex, age_band)

# Yearly table by sex and definition (aggregated over age bands)
incidence_table_year_sex <- incidence_data %>%
  group_by(year, sex_at_instance_creation, asthma_type) %>%
  summarise(
    events = sum(events),
    total_person_years = sum(total_person_years),
    incidence_per_1000py = (events / total_person_years) * 1000,
    .groups = "drop"
  ) %>%
  mutate(
    sex = recode(sex_at_instance_creation, "F" = "Female", "M" = "Male"),
    total_person_years = round(total_person_years, 2),
    incidence_per_1000py = round(incidence_per_1000py, 3)
  ) %>%
  select(year, sex, asthma_type, events, total_person_years, incidence_per_1000py) %>%
  arrange(asthma_type, year, sex)

# Overall table (2010-2022 pooled) by sex and definition
incidence_table_overall <- incidence_data %>%
  group_by(sex_at_instance_creation, asthma_type) %>%
  summarise(
    events = sum(events),
    total_person_years = sum(total_person_years),
    incidence_per_1000py = (events / total_person_years) * 1000,
    .groups = "drop"
  ) %>%
  mutate(
    sex = recode(sex_at_instance_creation, "F" = "Female", "M" = "Male"),
    total_person_years = round(total_person_years, 2),
    incidence_per_1000py = round(incidence_per_1000py, 3)
  ) %>%
  select(sex, asthma_type, events, total_person_years, incidence_per_1000py) %>%
  arrange(asthma_type, sex)

# Preview the tables that can be inserted in the report
knitr::kable(
  incidence_table_overall,
  caption = "Overall incidence rates (2010-2022 pooled), by sex and asthma definition"
)

knitr::kable(
  incidence_table_year_sex,
  caption = "Yearly incidence rates by sex and asthma definition"
)

rmarkdown::paged_table(incidence_table_stratum)
```

```{r save_incidence_tables, echo=TRUE}
dir.create("tables", showWarnings = FALSE, recursive = TRUE)

write_csv(incidence_table_stratum, "tables/incidence_table_stratum.csv")
write_csv(incidence_table_year_sex, "tables/incidence_table_year_sex.csv")
write_csv(incidence_table_overall, "tables/incidence_table_overall.csv")
```

## 6. Visualize Results

### 6.1 Create Plots

```{r prep_plot_data}
# Consistent labels/order for plots
age_levels <- c("0-19", "20-39", "40-59", "60+")

incidence_data <- incidence_data %>%
  mutate(
    age_band = factor(age_band, levels = age_levels),
    sex = recode(sex_at_instance_creation, "F" = "Female", "M" = "Male")
  )

# Yearly trend by sex, split by definition
plot_data_trend_sex <- incidence_data %>%
  group_by(year, sex, asthma_type) %>%
  summarise(
    events = sum(events),
    total_person_years = sum(total_person_years),
    incidence_per_1000py = (events / total_person_years) * 1000,
    .groups = "drop"
  )

# Age-band comparison by sex, pooled over 2010-2022
plot_data_age_band <- incidence_data %>%
  group_by(age_band, sex, asthma_type) %>%
  summarise(
    events = sum(events),
    total_person_years = sum(total_person_years),
    incidence_per_1000py = (events / total_person_years) * 1000,
    .groups = "drop"
  )

# Heatmap data (year x age-band), stratified by sex and definition
plot_data_heatmap <- incidence_data %>%
  group_by(year, age_band, sex, asthma_type) %>%
  summarise(
    events = sum(events),
    total_person_years = sum(total_person_years),
    incidence_per_1000py = (events / total_person_years) * 1000,
    .groups = "drop"
  )

# Possible vs narrow overall trend
plot_data_definition <- incidence_data %>%
  group_by(year, asthma_type) %>%
  summarise(
    events = sum(events),
    total_person_years = sum(total_person_years),
    incidence_per_1000py = (events / total_person_years) * 1000,
    .groups = "drop"
  )
```

#### 6.1.1 Time trends

```{r plot1_yearly_trend_by_sex, fig.width=10, fig.height=7}
plot1 <- ggplot(plot_data_trend_sex, aes(x = year, y = incidence_per_1000py, color = sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.7) +
  facet_wrap(~asthma_type, ncol = 1) +
  scale_color_manual(values = c("Female" = "#c44e52", "Male" = "#4c72b0")) +
  scale_x_continuous(breaks = seq(2010, 2022, by = 1)) +
  labs(
    title = "Yearly Incidence Trend by Sex",
    x = "Year",
    y = "Incidence rate per 1,000 person-years",
    color = "Sex"
  ) +
  theme_minimal()

plot1
```

#### 6.1.2 Stratified Rates

```{r plot2_ageband_comparison, fig.width=10, fig.height=7}
plot2 <- ggplot(plot_data_age_band, aes(x = age_band, y = incidence_per_1000py, fill = sex)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  facet_wrap(~asthma_type, ncol = 1) +
  scale_fill_manual(values = c("Female" = "#c44e52", "Male" = "#4c72b0")) +
  labs(
    title = "Age-Band Incidence Comparison by Sex (Pooled 2010-2022)",
    x = "Age band",
    y = "Incidence rate per 1,000 person-years",
    fill = "Sex"
  ) +
  theme_minimal()

plot2
```

```{r plot3_year_age_heatmap, fig.width=11, fig.height=8}
plot3 <- ggplot(plot_data_heatmap, aes(x = year, y = age_band, fill = incidence_per_1000py)) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_grid(asthma_type ~ sex) +
  scale_x_continuous(breaks = seq(2010, 2022, by = 2)) +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b") +
  labs(
    title = "Year x Age-Band Heatmap of Incidence Rates",
    x = "Year",
    y = "Age band",
    fill = "IR / 1,000 PY"
  ) +
  theme_minimal()

plot3
```

```{r plot4_possible_vs_narrow, fig.width=10, fig.height=6}
plot4 <- ggplot(plot_data_definition, aes(x = year, y = incidence_per_1000py, color = asthma_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c("Asthma_possible" = "#1b9e77", "Asthma_narrow" = "#d95f02")) +
  scale_x_continuous(breaks = seq(2010, 2022, by = 1)) +
  labs(
    title = "Possible vs Narrow Definition Comparison Over Time",
    x = "Year",
    y = "Incidence rate per 1,000 person-years",
    color = "Definition"
  ) +
  theme_minimal()

plot4
```

### 6.2 Key Findings

```{r save_plots, echo=TRUE}
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggsave("figures/figure1_yearly_trend_by_sex.png", plot = plot1, width = 10, height = 7, dpi = 300)
ggsave("figures/figure2_ageband_comparison_by_sex.png", plot = plot2, width = 10, height = 7, dpi = 300)
ggsave("figures/figure3_year_age_heatmap.png", plot = plot3, width = 11, height = 8, dpi = 300)
ggsave("figures/figure4_possible_vs_narrow_trend.png", plot = plot4, width = 10, height = 6, dpi = 300)
```
